A Systematic Literature Review on Extensions to Role Based Access Control

### 2013-03-24 paper/sections/abstract.tex
Breaks each research quesiton into it's own major section.

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions called roles.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: \textit{The goal of this work is to provide practioners an assessment of the state of extended models of RBAC and researchers with insights into the lack of robust evaluation of RBAC extension models.}
\\
\\
\noindent
{\bf Method}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}:
\\
\\
\noindent
{\bf Conclusions}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. We initially collect XXXX papers from sources such as IEEE and ACM 
websites and selected X primary sources.  We performed a comparative analysis of the primary sources to 
find relationships among extended models and analyze the state of RBAC extension evaluations.

>>>>

\noindent
{\bf Context}: Since the introduction of the role based access control (RBAC) standard in the late 1990's, 
RBAC has become an increasingly popular access control mechanism for applications such as 
web applications, mobile computing and databases. RBAC restricts access to resources 
based on the identity of subjects connected to logical groupings of permissions.  
The introduction of computing into new domains has led to proposals for extensions to the standard
RBAC model; for example, defining additional constraints among roles, adding context or defining new heirarchy relationships.
\\
\\
\noindent
{\bf Aim}: Our work aims to aid researchers looking to expand on the body of knowledge surrounding extensions to RBAC need a starting place to prevent re-invention of the wheel and
developers needing to find and select an access control mode based on their project requirements. 
\textit{The goal of this work is to provide practioners an assessment of the current extensions to RBAC to aid with choosing a model that meets their requirements, and researchers with insights into the state of RBAC extension models and how they are evaluated.}
\\
\\
\noindent
{\bf Method}: We conducted a systematic literature review by collecting and synthesizing relevant research
papers in the area of RBAC extensions. The review we performed yielded 1716 papers, of which 29 were deemed primary sources 
for inclusion as model extensions to the RBAC standard model. The papers we collected were from electronic digital libraries
IEEE, Google Scholar, CiteSeerX and ACM based on a set of exclusion and inclusion criteria.  
We performed a comparative analysis of the primary sources to find relationships among extended models and to analyze the state of RBAC extension evaluations.
\\
\\
\noindent
{\bf Results}: Our results showed that extensions to RBAC fall into a number of categories that in turn all
fall under the single category of context based extensions.  We look into the motivations behind RBAC extensions and
the domains that are leading factors in developing these newer models. We also quantify the current evaluation methods
used when RBAC extension models are presented to the research community.
\\
\\
\noindent
{\bf Conclusions}: The extensions to RBAC fall into a number of categorizations with Organization, Privacy, Resource, Task, Spatio-Temporal, Spatial and Temporal falling under the general category of context.
Domains, such as healthcare and mobile computing, were idenitfied as motivations behind the development of extensions to the RBAC model.  Our literature review showed that the state of RBAC model evaluation could benefit from the research community given most model evaluations seen within the papers were based on hypothetical situations with little to no case studies or implementations in practice.

### 2013-03-24 paper/sections/analysis.tex
Breaks each research quesiton into it's own major section.

\section{Analysis} \label{sec:analysis}

@@

\begin{itemize}
\item Core RBAC model needs additional contextual information and constraints to develop fine-grained policies in practice.
\item Core RBAC does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item Core RBAC does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item Core RBAC does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item Core RBAC has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item Model needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item Model needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
are introduced. 
\item Model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

@@

Of the 29 papers surveyed, there was a significant lack of implementation with 21 of the paper providing no
mention of a implementation or prototype.  Of the remaining 8 papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

@@

\begin{itemize}
\item Time-based Performance \cite{ni2010privacy}, \cite{aich09:role}
\item Complexity analysis \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{chen08:spatio-temporal}, \cite{aich09:role}
\item Comparison to standard RBAC \cite{bao08:role}, \cite{zou2009crbac}, \cite{zhang06:collaborative}, \cite{zhao2008flexible}, \cite{ray07:spatio}
\item Mathematical modeling \cite{damian2007geo}, \cite{hansen2003spatial}, \cite{aich07:STARBAC}, \cite{chen08:spatio-temporal}, \cite{joshi05:generalized}
\item Example scenarios of the model in action \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{cholewka00:acontext-sensitive}, \cite{huang06:pervasive}, \cite{bao08:role}, \cite{jian2008extended}, \cite{yamazaki104:designing}, \cite{zou2009crbac}, \cite{ray07:spatio}, \cite{samuel07:spatio-temporal}, \cite{ray07:spatio}, \cite{joshi05:generalized}, \cite{yao2008task}, \cite{zhou2007team}, \cite{oh2003task}
\item Experimental analysis of the model
\item Case study of the model in practice \cite{motta03:contextual}
\end{itemize}

@@

TODO: Add references to list above.

@@

\begin{itemize}
\item What domains or scenarios serve as inspiration for these extensions?
\item Medical domain
\item Pervasive computing environments
\item Mobile devices
\item Large-scale organizations with many sub-departments
\item Enterprise, organization workflows
\end{itemize}

>>>>

\section{Motivations} \label{sec:motivations}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Core RBAC model needs additional contextual information and constraints to develop fine-grained policies in practice.
\item Core RBAC does not incorporate context. Therefore, RBAC belongs to static access control model, which may not capture changes in environments.
\item Core RBAC does not support for various constraints such as temporal and spatial constraints to design sophisticated policies on demand.
\item Core RBAC does not provide an abstraction to additional user-defined attributes	(e.g., task and team) and their association with existing attributes.
\item Core RBAC has limitation on delegation and role hierarchy. For example, partial inheritance in role hierarchy needs to be developed.  
\item Model needs to incorporate additional contextual information and constraints to develop fine-grained policies in practice.
\item Model needs to incorporate additional attributes such as
task, team, purpose, organizational roles and collaborative activities over existing attributes. Moreover, new association between attributes
are introduced. 
\item Model needs to improve existing features such delegation and role hierarchy to meet fine-grained access control. 
\end{itemize}

\section{Implementations} \label{sec:implementations}

@@

Of the 29 papers surveyed, there was a lack of implementation with 21 of the paper providing no
mention of a implementation or prototype.  Of the remaining 8 papers that did mention an implementation, half 
were simply prototypes developed by the authors while the other half were claimed to be implemented within a real
system.

\section{Evaluations} \label{sec:evaluations}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Time-based Performance \cite{ni2010privacy}, \cite{aich09:role}
\item Complexity analysis \cite{bao08:role}, \cite{zhang06:collaborative}, \cite{chen08:spatio-temporal}, \cite{aich09:role}
\item Comparison to standard RBAC \cite{bao08:role}, \cite{zou2009crbac}, \cite{zhang06:collaborative}, \cite{zhao2008flexible}, \cite{ray07:spatio}
\item Mathematical modeling \cite{damian2007geo}, \cite{hansen2003spatial}, \cite{aich07:STARBAC}, \cite{chen08:spatio-temporal}, \cite{joshi05:generalized}
\item Example scenarios of the model in action \cite{alam06:constraint}, \cite{tzelepi01:flexible}, \cite{cholewka00:acontext-sensitive}, \cite{huang06:pervasive}, \cite{bao08:role}, \cite{jian2008extended}, \cite{yamazaki104:designing}, \cite{zou2009crbac}, \cite{ray07:spatio}, \cite{samuel07:spatio-temporal}, \cite{ray07:spatio}, \cite{joshi05:generalized}, \cite{yao2008task}, \cite{zhou2007team}, \cite{oh2003task}
\item Experimental analysis of the model
\item Case study of the model in practice \cite{motta03:contextual}
\end{itemize}

@@

\section{Domains} \label{sec:domains}

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item What domains or scenarios serve as inspiration for these extensions?
\item Medical domain
\item Pervasive computing environments
\item Mobile devices
\item Large-scale organizations with many sub-departments
\item Enterprise, organization workflows
\end{itemize}

\section{Generaliations} \label{sec:generalizations}

### 2013-03-24 paper/sections/appendix.tex
Breaks each research quesiton into it's own major section.

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{A constraint based role based access control in the SECTET a model-driven approach, 2006 \cite{alam06:constraint}} \\ \cline{2-4}

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{A flexible content and context-based access control model for multimedia medical image database systems, 2001 \cite{tzelepi01:flexible}} \\

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{Designing an agent-based RBAC system for dynamic security policy, 2004 \cite{yamazaki104:designing}} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Spatial }}}} &
\multicolumn{1}{ |l| }{GEO-RBAC: a spatially aware RBAC, 2005 \cite{damian2007geo}} \\[0.25ex] \cline{4-4}

>>>>

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{ A Context-Aware Role-Based Access Control Model for Web Services, 2005 \cite{haibo2005context}} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{}  &
\multicolumn{3}{ |l| }{Designing an agent-based RBAC system for dynamic security policy, 2004 \cite{yamazaki04:designing}} \\ \cline{2-4}

@@

\multicolumn{1}{ c }{} & \multicolumn{1}{ |c }{} & \multicolumn{1}{ |c }{{\multirow{3}{*}{\rotatebox{90}{ Spatial }}}} &
\multicolumn{1}{ |l| }{GEO-RBAC: a spatially aware RBAC, 2005 \cite{damiani2007geo}} \\[0.25ex] \cline{4-4}

### 2013-03-24 paper/sections/categorization.tex
Breaks each research quesiton into it's own major section.

@@

After deciding upon a set of primary sources, the authors noticed that each paper fell into one or more categorization's of extensions to RBAC.
The categorizations were extracted by looking at how they are defined, what are their characteristics and how do individual categories relate
to other categories, and encapsulated by the following research question:

What categorizations exist within RBAC extensions? \\

\subsection{Definitions}

@@

\subsection{Relationships}

>>>>

\section{Categorization} \label{sec:categorization}

After selection of primary sources, the next step was to extract data from each paper that pertained to our research questions in order to look for trends.  The first step was to take the individual data generated from the final elimination round and organize this information around the research questions.  During the paper reading round, and resulting data, the fact that the papers were logically falling into a number of categorizations became evident.  Thus, the first step undertaken was to answer the question of what categories exist for the RBAC model extensions and what papers feel into what categories.  Each paper deemed a primary source and the papers associated categorization are shown in Table \ref{tab:categorization}.

@@

What categories exist within RBAC extensions? \\

@@

### 2013-03-24 paper/sections/conclusion.tex
Breaks each research quesiton into it's own major section.

P1: Mention core RBAC, note specific results for each research question

P2: Mention bigger results that cut across and any other peritentn information

>>>>

The extensions to RBAC were revealed to fall into a number of categorizations with Organization, Privacy, Resource, Task, Spatio-Temporal, Spatial and Temporal falling under the general category of context.
The categories each had properties specific to their implementation, but were seen to generalize to being specialized instances of context tailored to the entities or actions the categories covered.
A number of domains were idenitfied as being the motivations behind needing extensions to the core RBAC model.  The domains, such as healthcare, presenting new challenges the previous models were not
required to design for.  Our literature review showed that the state of RBAC model evaluation needs focused from the research community given most model evaluations seen within the papers were
based on hypothetical situations with little to no case studies or implementations in practice.

### 2013-03-24 paper/sections/core_rbac.tex
Breaks each research quesiton into it's own major section.

\section{Core Role Based Access Control} \label{sec:core-rbac}

Since the basis for our review is extensions to the core model, we will describe the core model, associated entities and other terminology encountered across the space of our review.  The NIST RBAC model proposed by Ferraiolo et al. and later adopted as the official standard for RBAC by the International Committee for Information Technology Standards (INCITS) consists of four basic entities:

\begin{itemize}
\item a set of users \emph{Users}: A user can be a person or an agent.
\item  a set of roles \emph{Roles}: A role is a collection of permissions to perform a specific job function in an organization.
\item a set of permissions \emph{Permissions}: A permission refers to an access mode that can be exercised on an object in the system and a session relates a user to possibly many roles.
\item a set of sessions \emph{Sessions}: In each session, a user can be assigned to some of the roles, only when the corresponding role is enabled for activation for that time.		
\end{itemize}

In the RBAC, a user can exercise a permission only if the user are assigned to a role.
In addition to the four basic components, two functions are defined:
the user role assignment (UA) and the role
permission assignment (PA) functions.
UA models assignment of users to roles.
PA models assignment of permissions to roles.

>>>>

\section{Role Based Access Control Standard} \label{sec:core-rbac}

Since the basis for our review is extensions to the core model, we will describe the core model, associated entities and other terminology encountered across the space of our review.  The NIST RBAC model proposed by Ferraiolo et al. \cite{ferraiolokuhn} and later adopted as the official standard for RBAC by the International Committee for Information Technology Standards (INCITS) consists of four basic entities:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item a set of \emph{Users}: A user can be a person or an agent.
\item a set of \emph{Roles}: A role is a collection of permissions to perform a specific job function in an organization.
\item a set of \emph{Permissions}: A permission refers to an access mode that can be exercised on an object in the system and a session relates a user to possibly many roles.
\item a set of \emph{Sessions}: In each session, a user can be assigned to some of the roles, only when the corresponding role is enabled for activation for that time.		
\end{itemize}

In RBAC, a user can exercise a permission only if the user is assigned to a role.
In addition to the four basic components, two functions are defined:
the user role assignment (UA) and the role
permission assignment (PA) functions.
UA models assignment of users to roles.
PA models assignment of permissions to roles.

### 2013-03-24 paper/sections/intro.tex
Adds heirarchy table in appendix. Adds structured abstract.
Commiting updated sections from JeeHyun and a png version of the core model.
Breaks each research quesiton into it's own major section.

Role based access control (RBAC) was first introduced in the 1990s when the National Institute for Standards and Technology (NIST) requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework proposed by Sandhu, et al \cite{sandhu1996role}.  In an RBAC model, roles represent a group of users who are involved in a specific job function in an organization. RBAC assigns permissions of specific actions on resources to roles instead of individual users.  Therefore, in order to gain roles' permission on specific resources, users acquire appropriate roles first.  RBAC is a generalized access control approach used for various applications including web services, database applications, and healthcare applications.  RBAC has advantages in maintaining and managing organization's security policies.  For example, if a user is to access manager role's resources within a given organization, security policy administrators simply add the user to be associated with the manager role.

Since the introduction of the standardized RBAC model, innovation and the spread of software engineering into new domains has led to scenarios that, by some accounts, the standard RBAC model cannot handle.  For example, RBAC is ill-equipped to handle the additional entities that need to be taken into consideration during authorization with the introduction of privacy concerns in domains such as healthcare  Thus, a series of extensions to the standard RBAC model have appeared in the literature, each adding one or more features.  However, as researchers often develop their own specialized extended models of RBAC, their research cannot be generalized or compared with other research work appropriately.

\textit{The goal of this work is to provide practitioners an assessment of the state of extended models of RBAC and researchers with insights into the lack of robust evaluation of RBAC extension models.} With respect to our goal, the authors addressed the following research questions:

\begin{itemize}
\item What categorizations exist within extensions to RBAC?
\item What are the motivations behind extensions to RBAC?
\item Do the extensions to RBAC have corresponding implementations?
\item How are extensions to RBAC evaluated theoritically and in practice?
\item What domains have extensions to RBAC been created for?
\item What commonalities or generalizations exist across all categorizations?
\end{itemize}

A systematic literature review is designed to yield quantifiable results across a body of research in order to draw out similarities, trends and deficiencies in an area as a whole.  The review we performed yielded 1716, of which XX were deemed primary sources for inclusion as model extensions to the RBAC standard model.  This review is intended to serve as a starting place for researcher's looking to tackle new problems in the realm of authorization and prevent re-invention of the wheel. For developer's looking to find a model to fit their seemingly unique access control needs, this review will provide a basis for comparison and easy look-up for what extension based model to use.  Further, the review provides insight into the state of evaluation, or lack thereof, within the RBAC extension model community.

@@

\begin{itemize}
\item Summarizes current extended RBAC research work and its contributions
\item Guides a direction for a standard of extended RBAC. Understanding the categorization and the motivation of the existing research results helps decide a standard of extended RBAC.
\item Our work shows a criteria in comparison among research results.
\item Identify the research challenges in the areas of security policies and suggest a future extension of RBAC
\item Identify the deficiencies in RBAC extension evaluation
\end{itemize}

The rest of the paper is organized as follows.  Section 2 covers background and the core RBAC model. Section 3 lays out the process used in conducting the review. Section 4 provides the results of the search. Section 5 tackles the research questions and analyzes the results.  Section 6 contains the final conclusions.

\begin{itemize}
\item TODO: Add some references to the prevalence of RBAC in industry
\item TODO: Add link to RBAC standard documents
\item TODO: List research questions here or in process
\end{itemize}

>>>>

Innovation and the use of RBAC into new domains has led to scenarios that, by some accounts, the standard RBAC model cannot handle.  For example, Ni, Q. and Bertino, E. et al \cite{ni2010privacy} state that RBAC is ``not designed to enforce privacy policies and barely meet privacy protection requirements`` with the introduction of privacy concerns in domains such as healthcare. Thus, a series of extensions to the standard RBAC model have appeared in the literature, each adding one or more features.  However, as researchers often develop their own specialized extended models of RBAC, their research cannot be generalized or compared with other research work appropriately.

Role based access control (RBAC) was first introduced in the 1990s when the National Institute for Standards and Technology (NIST) requested that a unified standard be created by combining the Ferraiolo and Kuhn model \cite{ferraiolokuhn} with the framework proposed by Sandhu, et al \cite{sandhu1996role}.  The development of a standard was inspired by an economic impact study done during the 1990s and later confirmed in 2011 \footnote{\url{http://csrc.nist.gov/groups/SNS/rbac/documents/20101219_RBAC2_Final_Report.pdf}}. In an RBAC model, roles represent a group of users who are involved in a specific job function in an organization. RBAC assigns permissions of specific actions on resources to roles instead of individual users.  Therefore, in order to gain roles' permission on specific resources, users acquire appropriate roles first.  RBAC is a generalized access control approach used for various applications including web services, database applications, and healthcare applications.  RBAC has advantages in maintaining and managing organization's security policies.  For example, if a user is to access manager role's resources within a given organization, security policy administrators simply add the user to be associated with the manager role.

\textit{The goal of this work is to provide practioners an assessment of the current extensions to RBAC to aid with choosing a model that meets their requirements, and researchers with insights into the state of RBAC extension models and how they are evaluated.} With respect to our goal, the authors addressed the following research questions:

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item What categorizations exist within extensions to RBAC?
\item What are the motivations behind extensions to RBAC?
\item Do the extensions to RBAC have corresponding implementations?
\item How are extensions to RBAC evaluated theoritically and in practice?
\item What domains have extensions to RBAC been created for?
\item What commonalities or generalizations exist across all categorizations?
\end{itemize}

The authors performed a systematic literature review designed to explore the current body of research in the area of extensions to the core RBAC model.  The review we performed yielded 1716, of which 29 were deemed primary sources for inclusion as model extensions to the RBAC standard model.  This review is intended to serve as a starting place for researcher's looking to tackle new problems in the realm of authorization and prevent re-invention of the wheel. For developer's looking to find a model to fit their seemingly unique access control needs, this review will provide a basis for comparison and easy look-up for what extension based model to use.  Further, the review provides insight into the state of evaluation, or lack thereof, within the RBAC extension model community.

@@

\begin{itemize}
\setlength{\itemsep}{0.25pt}
\item Summarizes current extended RBAC research work and its contributions.
\item Guides a direction for a standard of extended RBAC. Understanding the categorization and the motivation of the existing research results helps decide a standard of extended RBAC.
\item Identifies the research challenges in the areas of security policies and suggest a future extension of RBAC.
\item Identifies the deficiencies in RBAC extension evaluation.
\end{itemize}

The rest of the paper is organized as follows.  Section 2 covers background and the core RBAC model. Section 3 lays out the process used in conducting the review. Section 4 discusses the categorization break down of the papers. Section 5 analyzes the rest of the research questions.  Section 6 contains the final conclusions.
